### 2021-06-11 abstractESSLLI21/abstractESSLLI1.Rmd
proofread abstract, small changes

\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}
A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as number vectors (see e.g. @bolukbasi2016man and @Gonen2019lipstick and references therein). The most common method used compares  cosine similarity between words from protected groups and attributes that are considered to be stereotypical or harmful in some way, and this method will be in our focus.

@@

Such methods are statistically problematic. One issue, specific to @Manzini2019blackToCriminal, is that no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word. A related, general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes. \todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"? albo jakoś podobnie} That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than 1.

Another, also serious problem, is that  all the authors calculate means of means and their authors \todo{jak rozumiec drugi raz powtorzone their authors?} run statistical tests on sets of means. Unfortunately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.

All such approaches come up with rather short lists of protected words and rather short lists of stereotypical attributes. Clearly, these are not complete list. So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously. To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation. 
\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}

@@

Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard deviation  (code available upon request).

>>>>

A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as  vectors of real numbers (see e.g. @bolukbasi2016man and @Gonen2019lipstick and references therein). The most common method used compares  cosine similarity between words from protected groups and attributes that are considered to be stereotypical or harmful in some way, and this method will be in our focus.

@@

Such methods are statistically problematic. Two issues, specific to @Manzini2019blackToCriminal, are that (i) no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word, and (ii) no re-scaling by standard deviation (typical in effect size measures such as  Cohen's $d$) is used, so MAC is not a good measure of effect size. A more general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.  That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than neutral non-human words.

Another, also serious problem, is that  all the authors calculate means of means and  run statistical tests on sets of means. Unfortunately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.

All such approaches come up with rather short lists of protected words and rather short lists of stereotypical attributes. Clearly, these are not complete list. So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously. To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation.

@@

We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).

### 2021-06-11 abstractESSLLI21/abstractESSLLI1.tex
proofread abstract, small changes

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
            colorlinks=true,
            linkcolor=Maroon,
            filecolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as number vectors (see e.g.~{[}1{]} and {[}4{]} and
references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

@@

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align} \normalsize \noindent {[}2{]} show that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. {[}5{]} extends the methodology to a
multilingual and cross-lingual setting; a similar methodology is
employed by {[}3{]}, who use word embeddings trained on corpora from
different decades. {[}6{]} modify WEAT to a multi-class setting,
introducing Mean Average Cosine (MAC) similarity as a measure of bias
(in fact, in the paper they report distances rather than similarities).
Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
and let each \(A_j\in A\) be a set of attributes stereotypically
associated with a protected word). Then:

@@

\begin{align}
\mathsf{S}(t_i, A_j)  = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
\end{align} \normalsize  \noindent That is, for each protected word and
each attribute class, they first take the mean for this protected word
and all attributes in a given attribute class, and then take the mean of
thus obtained means for all the protected words.

Such methods are statistically problematic. One issue, specific to
{[}6{]}, is that no distinction is made based on whether given a
protected word a class of attributes is stereotypically associated with
this protected word or with a different protected word. A related,
general problem, is that all the authors ignore the step of comparing
their results with control groups, especially control groups of
stereotype-neutral human attributes. That such a move is important is
suggested for instance by Figure 1 prepared using the original word list
for religion-related protected words extended with control attributes,
where such human attributes are also closer to protected classes than 1.

Another, also serious problem, is that all the authors calculate means
of means and their authors run statistical tests on sets of means.
Unforutnately, by pre-averaging the data they throw away information
about sample sizes, and they remove variation, and so pre-averaging
tends to manufacture false confidence.

@@

\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}
Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard devation  (code available upon request).
\end{minipage}
\end{tabular}

@@

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align} \normalsize

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
{[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
and Adam Kalai. 2016. Man is to computer programmer as woman is to
homemaker? Debiasing word embeddings. Retrieved from
\url{http://arxiv.org/abs/1607.06520}

\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
{[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
Semantics derived automatically from language corpora contain human-like
biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:\url{https://doi.org/10.1126/science.aal4230}

\leavevmode\hypertarget{ref-Garg2018years}{}%
{[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:\url{https://doi.org/10.1073/pnas.1720347115}

\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
{[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
methods cover up systematic gender biases in word embeddings but do not
remove them. In \emph{Proceedings of the 2019 conference of the north
American chapter of the association for computational linguistics: Human
language technologies, volume 1 (long and short papers)}, Association
for Computational Linguistics, Minneapolis, Minnesota, 609--614.
DOI:\url{https://doi.org/10.18653/v1/N19-1061}

\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
{[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
biased? Multidimensional analysis of biases in distributional word
vectors. \emph{CoRR} abs/1904.11783, (2019). Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
{[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
2019. Black is to criminal as caucasian is to police: Detecting and
removing multiclass bias in word embeddings. Retrieved from
\url{http://arxiv.org/abs/1904.04047}

\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
{[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.
2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:\url{https://doi.org/10.1037/1089-2699.6.1.101}

>>>>

\documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{
            colorlinks=true,
            linkcolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
and references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

@@

\begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}

\normalsize
\noindent {[}2{]} show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. {[}5{]} extends the methodology to a multilingual and
cross-lingual setting; a similar methodology is employed by {[}3{]}, who
use word embeddings trained on corpora from different decades. {[}6{]}
modify WEAT to a multi-class setting, introducing Mean Average Cosine
(MAC) similarity as a measure of bias (in fact, in the paper they report
distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
a class of protected word embeddings, and let each \(A_j\in A\) be a set
of attributes stereotypically associated with a protected word). Then:

@@

\begin{align}
\mathsf{S}(t_i, A_j)  = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
\end{align}

\normalsize  \noindent That is, for each protected word and each
attribute class, they first take the mean for this protected word and
all attributes in a given attribute class, and then take the mean of
thus obtained means for all the protected words.

Such methods are statistically problematic. Two issues, specific to
{[}6{]}, are that (i) no distinction is made based on whether given a
protected word and a class of attributes is stereotypically associated
with this protected word or a different protected word, and (ii) no
re-scaling by standard deviation (typical in effect size measures such
as Cohen's \(d\)) is used, so MAC is not a good measure of effect size.
A more general problem, is that all the authors ignore the step of
comparing their results with control groups, especially control groups
of stereotype-neutral human attributes. That such a move is important is
suggested for instance by Figure 1 prepared using the original word list
for religion-related protected words extended with control attributes,
where such human attributes are also closer to protected classes than
neutral non-human words.

Another, also serious problem, is that all the authors calculate means
of means and run statistical tests on sets of means. Unfortunately, by
pre-averaging the data they throw away information about sample sizes,
and they remove variation, and so pre-averaging tends to manufacture
false confidence.

@@

\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}

We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).

\end{minipage}
\end{tabular}

@@

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align}

\normalsize

@@

\hypertarget{refs}{}
\hypertarget{ref-bolukbasi2016man}{}
{[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
and Adam Kalai. 2016. Man is to computer programmer as woman is to
homemaker? Debiasing word embeddings.

\hypertarget{ref-Caliskan2017semanticsBiases}{}
{[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
Semantics derived automatically from language corpora contain human-like
biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:\url{https://doi.org/10.1126/science.aal4230}

\hypertarget{ref-Garg2018years}{}
{[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:\url{https://doi.org/10.1073/pnas.1720347115}

\hypertarget{ref-Gonen2019lipstick}{}
{[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
methods cover up systematic gender biases in word embeddings but do not
remove them. In \emph{Proceedings of the 2019 conference of the north
American chapter of the association for computational linguistics: Human
language technologies, volume 1 (long and short papers)}, Association
for Computational Linguistics, Minneapolis, Minnesota, 609--614.
DOI:\url{https://doi.org/10.18653/v1/N19-1061}

\hypertarget{ref-Lauscher2019multidimensional}{}
{[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
biased? Multidimensional analysis of biases in distributional word
vectors. \emph{CoRR} abs/1904.11783, (2019). Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\hypertarget{ref-Manzini2019blackToCriminal}{}
{[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
2019. Black is to criminal as caucasian is to police: Detecting and
removing multiclass bias in word embeddings.

\hypertarget{ref-Nosek2002harvesting}{}
{[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.
2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:\url{https://doi.org/10.1037/1089-2699.6.1.101}
